A template for the \emph{arxiv} style

### 2022-12-14 paper-rmd/paper-papaja-revision2.Rmd
updated rmd and avg code

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic, but people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These models differ in their averaged predictions about people's conditional probability judgments and in their distributional predictions about their overall patterns of judgments. In particular, the Bayesian sampler's Bayesian adjustment process predicts a truncated range of responses as well as a correlation between the average degree of bias and variability trial-to-trial. However, exploring these distributional predictions with participants' raw responses requires a careful treatment of rounding errors and exogenous response processes. Here, I cast these theories into a Bayesian data analysis framework that supports the treatment of these issues along with principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by an account of biases based on "noise" in the sample-reading process.

@@

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "9374"

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
biblio-style      : "apa"
documentclass     : "apa6"
classoption       : "man"
output:
  papaja::apa6_word
keep_tex: yes
always_allow_html: true
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{bayesnet}
  - \mathchardef\hyph="2D
---

@@

Trouble starts as soon as we attempt to measure beliefs. According to Bayesian theories of cognition and epistemology [@jaynes2003], the degree to which people believe in various propositions, or their credences, should reflect subjective mental probabilities. So asking people to express beliefs in terms of probability seems only natural.

Unfortunately, people’s explicit probability judgments routinely violate the most basic axioms of probability theory. For example, human probability judgments often exhibit the "conjunction fallacy": people will often judge the conjunction of two events (e.g. "Tom Brady likes football and miniature horses") as being more probable than one of the events in isolation (e.g. "Tom Brady likes miniature horses"), a plain and flagrant violation of probability theory [@tversky.kahneman1983]. Other demonstrations of the incoherence of probability judgments include disjunction fallacies, subadditivity or "unpacking" effects [@tversky.koehler1994], and a number of others [for an accessible review, see @kahneman2013]. Altogether these findings have led many researchers to abandon the notion that degrees of belief are represented as probabilities.

@@

Costello and Watts [-@costello.watts2014; -@costello.watts2016; -@costello.watts2018] have proposed a theory of probability judgment they call the "Probability Theory plus Noise" theory (PT+N). In the PT+N model, mental "samples" are drawn from a probabilistic mental model of events and are then "read" with noise, so that some positive examples will be read as negative and some negative examples read as positive with some probability $d$. The end products are probability judgments reflecting probabilistic credences perturbed by noise. In their model, the probability a mental sample for an event $A$ is read as $A$ is the probability that the sample truly is $A$, $p(A)$, and that it is correctly read $(1-d)$, plus the probability that the sample is not $A$, $1-P(A)$ and that it is incorrectly read ($d$), or:

@@

Thus, in many cases the effect of a Bayesian prior is identical to the effect of noise in the PT+N model (at least in expectation). But, rather than merely perturbing people's probability judgments, this prior can be seen as regularizing these judgments away from extreme values. Zhu and colleagues [-@zhu.etal2020] argue that such regularization can be adaptive in cases where only a small number of mental samples can be drawn. For instance, consider someone estimating the probability that they can swim across a lake, outrun an animal, or win a hand of poker: if a mental simulation of these events produces two samples indicating success, one might conclude these are all certain victories and thereby be too willing to assume risk. A regularizing prior pushes these estimates away from extremes, thereby promoting better decision-making when mental samples are sparse. However, this hedging comes at the cost of systematic incoherence and biases.

@@

The model's predictions can be distinguished on two levels: First, the models have distinct accounts of conditional probability judgments that make different predictions in terms of expected values. Second, the models present different process-level accounts of probability judgment that entail different predictions about the shape of the distribution of responses across individuals.

@@

In later work, Sundh et al [-@sundh.etal2021] examined the distributional properties of participants responses using indirect means, by regressing the variance of participants' responses across trials on their mean response for each query type. Their findings suggest that earlier fits with Wasserstein distance may have produced biased results [@sundh.etal2021]. They reported evidence for the truncation of responses and a correlation between variance and shrinkage parameter estimates across participants. However, their analysis did not enforce that the underlying probabilities driving participants' judgments be coherent (simply estimating the true probability as the mean across trials), nor did they evaluate how frequently participants gave out-of-support judgments that would be inconsistent with the Bayesian Sampler theory.

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. <<revise: add that I will take two approaches, first examining their predictions in expectation to test their different accounts of conditional probability judgments. Then testing their distributional predictions to test their process-level accounts of probability judgments >>

The Bayesian framework supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in potential improvements to out-of-sample prediction, reductions in model complexity, and a more realistic test of the models. In addition, Bayesian data analysis allows issues of model complexity to be addressed through comparisons of model fit based on modern information criteria, such as Pareto smoothed importance sampling approximate leave-one-out cross validation ($\text{PSIS-LOO}$; [@gelman.etal2014; @vehtari.etal2017]).[^2]

@@

Finally, I implement new extensions of these models to directly model participants trial-level responses while accounting for rounding and out-of-distribution response errors---allowing for principled probabilistic tests of the distributional predictions of the models.

@@

<< this first set of analyses compares the models' ability to capture participant's proability judgments in expectation, averaged over the three blocks on which they made judgments about each query. These analyses will test the first points of differentiation between the models, their different predictions with respect to conditional probability judgmetns and their specific parameterizations.>>

@@

Both of these models can also be implemented as hierarchical models with partial pooling for the $d$ and $d'$ parameters (implicitly, for $N$ and $N'$ in the case of the Bayesian Sampler). This partial pooling can help to regularize parameter estimates and improve out-of-sample predictive performance. In addition, partial pooling effectively reduces model complexity, and could support more realistic comparison between the "simple" and "complex" variants of the models.Figure 2 displays the translation of a hierarchical implementation of the Bayesian Sampler model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters. For ease of interpretation, the centered parameterization is shown below, although the actual models used a non-centered parameterization to improve sampling efficiency [@papaspiliopoulos.etal2007].

@@

Finally, I also explored fitting a hierarchical version of the Bayesian Sampler model that allowed values of $\beta > 1$. Restricting $\beta$ to [0,1] restricts the prior distribution of the Bayesian sampler to the class of "ignorance priors" [@zhu.etal2020]. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted [see e.g., @fennell.baddeley2012]. If $\beta$ is unrestricted, allowed to fall in the domain $[0, \infty]$ then the Bayesian Sampler model becomes more flexible, allowing for equivalent "noise" levels in the same $[0, 1/2]$ range as the PT+N model. That is, through the bridging condition, the implied $d$ approaches $1/2$ in the limit as $N \to 1$ and $\beta \to \infty$. Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

Data from Experiment 1 favor "complex" variants of the Bayesian Sampler model compared with the "simple" variants and all versions of the PT+N model (greater values of $\widehat{\text{elpd}}_{\text{LOO}}$ are better). However, there is no clear single winner. As shown in Figure 3, the best-scoring model is an unrestricted variant of the Bayesian Sampler that allows for people to bring informative priors to the probability judgment task (i.e. allowing $\beta \in [0, \infty]$. However, the difference between this and the next best-fitting models are within two standard errors of the difference, indicating that these models are also plausible [@sivula.etal2020]. Data from Experiment 2 more decisively reveal a single winning model: the hierarchical "unrestricted" implementation of the Bayesian Sampler model allowing for informative priors.

<< This unrestricted BS model differs from the PT+N model only in its treatment of conditional probability judgments. Thus, it can also be seen as a variant of the PT+N model that jettisons its constructive account of conditional probability judgments. From its superior fit, we can infer that the Bayesian Sample theory provides a better account of human conditional probability judgments.>>

@@

```{r, fig.cap="Posterior density of population-level $d$  and $d'$ parameters estimated from the unrestricted hierarchical Bayesian Sampler model for data from Experiments 1 and 2. Dashed line indicates theoretical maximum values for Bayesian Sampler model with uninformative priors."}
plt_d_avgmodel + plt_d_trialmodel + plot_layout(ncol=1, guides="collect") & labs(x="Estimate") & theme_bw() & theme(panel.grid=element_blank())

@@

t2_df %>% 
  select(-`...1`) %>% 
  gather(measure, value, -Experiment, -model) %>% 
  pivot_wider(names_from = c(Experiment, measure), values_from=value,names_sort = T) %>% 
  rename(Model = model) %>%
  kbl(
    caption="Bayesian model comparison results with best scoring model in bold face. Models are compared based on X and the correlation between their predictions for each participants' query-level average ($r_{\\text{query}}$) as well as overall query-level averages $r_{\\text{query-avg}}$.", 
    digits=3, 
    booktabs=T,
    col.names = c("Model", "$\\widehat{\\text{elpd}}_{\\text{LOO}}$", "$\\hat{p}_{\\text{LOO}}$","$r_{\\text{query}}$", "$r_{\\text{query-avg}}$", "$\\widehat{\\text{elpd}}_{\\text{LOO}}$", "$\\hat{p}_{\\text{LOO}}$","$r_{\\text{query}}$", "$r_{\\text{query-avg}}$"),
    escape = F
    ) %>%
  add_header_above(c(" ", "Experiment 1"=4, "Experiment 2"=4)) %>% 
  kable_classic(full_width=F) %>% 
  kable_styling(latex_options="scale_down") %>% 
  row_spec(1, bold=TRUE)
```

@@

```{r, fig.cap="Average model predicted and observed values for the 18 identities. Note that the Bayesian Sampler but not the PT+N model is capable of predicting non-zero values for identities Z10 through Z13. Error bars represent 95% CI. In Experiment 1, like participants' responses, the model's estimates are very slightly positive for \\{icy, frosty\\} and very slightly negative for \\{normal, typical\\}. This pattern replicates the qualitative pattern reported by Zhu and colleagues."}
plt_ineq_exp1 + labs(title="Exp. 1") + 
  plt_ineq_exp2 + labs(title="Exp. 2") + 
  plot_layout(guides="collect") & theme(legend.position="bottom") & labs(color="", y = "Estimate", x = "Identity")
```

@@

## Raw trial-level response distributions

The best-fitting model capturing participant's predictions in expectation can be seen either as a variant of the Bayesian Sampler theory allowing for informative priors or as a variant of the PT+N theory without any special treatment of conditional versus uncondtional probability judgments. That is to say, these prior analyses have not decisively ruled between the different process-level psychological theories behind the models. To address this question, response distributions must be considered.

Recall, in contrast to a noise-based account, the Bayesian Sampler theory predicts both a truncation of the response distribution as well as a correlation between the degree of shrinkage in probability estimates and the variability of those estimates trial-to-trial (assuming $\beta$ is relatively small). Because the Bayesian Sampler applies a Bayesian adjustment after sampling, it predicts probability judgments will always lie between $d$ and $1 - d$, even when zero positive or negative mental samples are drawn (see Equation 12 and the bridging condition, Equation 5). First, it bears noting that the truncated response distribution implied by the Bayesian Sampler model appears at odds with the raw response data. However, participants' responses frequently lie outside the range implied by the best estimates of their $d$ parameters (41% in Experiment 1 and 60% in Experiment 2).

@@

Under both models, the variability of people's responses trial-to-trial is driven by the number of mental samples drawn: more mental samples produce less-variable responses. However, if the number of samples is considered to be a truly discrete quantity, then only a limited number of discrete responses are possible. As Zhu and colleauges [-@zhu.etal2020] note, this is somewhat implausible on its face and their later work has abandoned this assumption [@zhu.etal2021].

At the same time, from a pragmatic perspective it is highly desirable that all latent parameters within the models be continuous. The models would be far more tractable to fit if, rather than including a latent Binomial variable representing the discrete number of samples drawn, we could instead model a continuous proportion of samples using, for instance, a Beta distribution.

Zhu and collauges [-@zhu.etal2020] introduce the possibility of an "autocorrelated" Bayesian Sampler model under which samples are assumed to be autocorrelated [ideas which were advanced further in [@zhu.etal2021]. As autocorrelated samples provide less information than i.i.d samples, they should be weighted when computing probability estimates. The idea is that people actually draw N autocorrelated samples that approximate some smaller $N_{\text{eff}}$ i.i.d samples. Assuming the true number of samples drawn, N, is allowed to vary somewhat noisily, then a model based on autocorrelated samples would no longer be limited to predicting a discrete set of possible responses.

<< say a bit more on this assumption / revise to clarify a bit >>

@@

First, it is clear that participants have rounded a majority of their responses. This sort of rounding can be modeled by a categorical distribution across the discrete possible rounded responses. Each rounded response category corresponds to a set of cut points, $a$ and $b$, with the probability of the categorical response defined by the cumulative distribution function of the underlying latent distribution.

@@

As participants were allowed to respond freely with whole numbers from 0 to 100, the exact rounding policy for each response is unknown. Nevertheless these rounding policies can be estimated via mixture modeling. For simplicity, rounding to the nearest 5% was enforced for all responses. Then, the probability of these categorical responses are computed for 21 and 11 categories (corresponding to rounding to 5% and 10%). These probabilities were combined along with a uniform probability representing "contaminants" according to mixing probabilities $\phi$, distributed with a Dirichlet prior (see Appendix for further implemenatation details).

@@

Above we found that an autocorrelated sampling process could be reasonably approximated by a Beta distribution. Taking the original Bayesian Sampler model,

@@

This model assumes that the number of samples drawn on each trial is fixed as $N$ or $N'$ accordingly (modulo the uncertainty about these parameters). However, it also seems reasonable to imagine that the number of samples drawn in fact varies from trial-to-trial. This added complexity is likely ignorable for a noise-based model, but could be important for the Bayesian Sampler model, as the number of samples drawn affects the truncation of the response distribution. That is, allowing the effective number of samples drawn to vary across trials may allow the model to capture some responses that would otherwise be treated as contaminants. To capture this, the model can be given one additional extension to allow the number of samples to vary, by adding a new parameter $N_{trial}$. This parameter multiplies the number of effective samples as a fraction of each individual participants' average number of samples drawn, e.g. so that a participant might sometimes draw 1.5x or 2x the number of effective samples they typically draw. The appropriate amount of variation in $N_{trial}$ is constrained to be fairly small, but is estimated hierarchically: I assume $log(N_{trial}) \sim N(0,\sigma_{N_{trial}})$ and $log(\sigma_{N_{trial}}) \sim (-1, .3)$.

@@

Prior to fitting the models, response data were rounded to the nearest 5%. Nearly all responses (Exp. 1: 93% and Exp. 2: 89%) were already divisible by five, and this was necessary to speed model fitting. Even still, estimating model posteriors for the trial-level mixture models using MCMC proved intractable. Instead, model posteriors were estimated using Stochastic Variational Inference (SVI) with Numpyro [@phan.etal2019] using a multivariate Normal guide [e.g. @kucukelbir.etal2015] and 10,000 SVI steps. Estimating each model posterior took about 40 minutes on an Nvidia V100 GPU. Simulation studies verified that this estimation approach could reliably recover parameters from simulated data (see Supplemental Materials).

@@

The non-varying Bayesian Sampler model fits quite poorly and is estimated to have a large proportion of "contaminant" responses (34% in Experiment 1, 25% in Experiment 2). This is likely due to a lack of the predicted truncation of the response distribution. As with the trial-average models, the estimates of implied d are quite high, which would predict substantial truncation. As mentioned earlier, many of participants' responses fall outside the range implied by their most likely implied d values.

Allowing the effective number of samples drawn to vary trial-by-trial improves the fit of the Bayesian Sampler model substantially and decreases the estimated proportion of contaminant responses (28% in Experiment 1, 12% in Experiment 2). Nevertheless, these results still indicate inferior fit compared with the noise-based model, which has substantially lower $\widehat{\text{elpd}}_{\text{LOO}}$ and again attributes fewer responses to the contaminant process (14% in Experiment 1, 4% in Experiment 2).

Second, we can examine the shrinkage-variance relationship. In the noise-based model, $N$ and $d$ were allowed to vary freely. But if these quantities are actually correlated as predicted by a Bayesian adjustment model, then we should expect to nevertheless see a correlation between the subject-level estimates in these parameters. Figure 11 shows a scatterplot relating these estimates. Looking across subjects we see they are mostly uncorrelated. Although there is a slight negative correlation in Experiment 1, this is driven by only a handful of participants with extreme values and in Experiment 2 the pattern appears to if anything be reversed. These findings again run counter to the predictions of the Bayesian Sampler theory.

```{r, fig.cap = "Scatterplots showing the relationships between subject-level d and N estimates from the trial-level noise-based model for Experiments 1 and 2. Figure for Experiment 1 excludes some outlier paricipants who gave repetitive resposnes that resulted in abnormally high N-values. Error bars indicate 95% CI.  "}

>>>>

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic, but people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [PT+N; @costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These models differ in their averaged predictions about people's conditional probability judgments and in their distributional predictions about their overall patterns of judgments. In particular, the Bayesian Sampler's Bayesian adjustment process predicts a truncated range of responses as well as a correlation between the average degree of bias and variability trial-to-trial. However, exploring these distributional predictions with participants' raw responses requires a careful treatment of rounding errors and exogenous response processes. Here, I cast these theories into a Bayesian data analysis framework that supports the treatment of these issues along with principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020], I find these data are best explained by an account of biases based on "noise" in the sample-reading process but in which conditional probability judgments are produced by a process of conditioning in the mental model of the events, rather than in a two-stage mental sampling process as proposed by the PT+N model.

@@

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "8484"

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
biblio-style      : "apa"
documentclass     : "apa6"
classoption       : "man"
output:
  papaja::apa6_pdf
keep_tex: yes
always_allow_html: true
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{bayesnet}
  - \mathchardef\hyph="2D
---

@@

Trouble starts as soon as we attempt to measure beliefs. According to Bayesian theories of cognition and epistemology [@jaynes2003], the degree to which people believe in various propositions, or their credences, should reflect subjective mental probabilities. So, asking people to express beliefs in terms of probability seems only natural.

Unfortunately, people’s explicit probability judgments routinely violate the most basic axioms of probability theory. For example, human probability judgments often exhibit the "conjunction fallacy": people will often judge the conjunction of two events (e.g. "Tom Brady likes football and miniature horses") as being more probable than one of the events in isolation (e.g. "Tom Brady likes miniature horses"), a plain and flagrant violation of probability theory [@tversky.kahneman1983]. Other demonstrations of the incoherence of probability judgments include disjunction fallacies, subadditivity or "unpacking" effects [@tversky.koehler1994], and a number of others [for an accessible review, see @kahneman2013]. Altogether, these findings have led many researchers to abandon the notion that degrees of belief are represented as probabilities.

@@

Costello and Watts [-@costello.watts2014; -@costello.watts2016; -@costello.watts2018] have proposed a theory of probability judgment they call the "Probability Theory plus Noise" theory (PT+N). In the PT+N model, mental "samples" are drawn from a probabilistic mental model of events and are then "read" with noise, so that some positive examples will be read as negative and some negative examples read as positive with some probability $d$. The end products are probability judgments reflecting probabilistic credences perturbed by noise. In their model, the probability that a mental sample for an event $A$ is correctly read as $A$ is the probability that the sample truly is $A$, $p(A)$, and that it is correctly read $(1-d)$, plus the probability that the sample is not $A$, $1-P(A)$ and that it is incorrectly read ($d$), or:

@@

Thus, in many cases the effect of a Bayesian prior is identical to the effect of noise in the PT+N model (at least in expectation). A caveat to this is that the Bayesian Sampler theory restricts the parameterization of the equivalent $d$ parameter compared to the PT+N model. Whereas PT+N assumes $d \in [0, 1/2]$, the Bayesian Sampler theory assumes an uninformatative prior parameter $\beta \in [0, 1]$, which in turn restricts the equivalent parameterization of $d \in [0, 1/3]$. But beyond this subtle difference of parameterization, there is a larger difference in interpretation: rather than merely perturbing people's probability judgments, this prior can be seen as regularizing these judgments away from extreme values. Zhu and colleagues [-@zhu.etal2020] argue that such regularization can be adaptive in cases where only a small number of mental samples can be drawn. For instance, consider someone estimating the probability that they can swim across a lake, outrun an animal, or win a hand of poker: if a mental simulation of these events produces two samples indicating success, one might conclude these are all certain victories and thereby be too willing to assume risk. A regularizing prior pushes these estimates away from extremes, thereby promoting better decision-making when mental samples are sparse. However, this hedging comes at the cost of systematic incoherence and biases.

@@

The model's predictions can be distinguished on two levels: First, the models have distinct accounts of conditional probability judgments that make different predictions in terms of expected values. Second, the models present different process-level accounts of probability judgment that entail different predictions about the shape of the distribution of responses across trials.

@@

In later work, Sundh et al [-@sundh.etal2021] examined the distributional properties of participants' responses using indirect means, by regressing the variance of participants' responses across trials on their mean response for each query type. Their findings suggest that earlier fits with Wasserstein distance may have produced biased results [@sundh.etal2021]. They reported evidence for the truncation of responses and a correlation between variance and shrinkage parameter estimates across participants. However, their analysis did not enforce that the underlying probabilities driving participants' judgments be coherent (simply estimating the true probability as the mean across trials), nor did they evaluate how frequently participants gave out-of-support judgments that would be inconsistent with the Bayesian Sampler theory.

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. Two sets of analyses compared different aspects of the models' predictions: First, the model's were examined based on their predictions in expectation. This analysis allows for a test of their different accounts of conditional probability judgments and their parameterizations (i.e. the restriction of $d$ to $[0, 1/2]$ versus $[0, 1/3]$). To preview the results, these analyses revealed the Bayesian Sampler's account of probability judgments to be superior, but could not distinguish between the different psychological processes proposed by the two models. A second set of analyses examined the model's distributional predictions to test their process-level accounts of probability judgments. As will be described, modeling response distributions directly presents a number of challenges, and so this second set of analyses required some minor additions and modifications to the models to permit their fitting to experimental data.

Both sets of analyses are supported by the use of a Bayesian Framework. First, Bayesian analyses allow issues of model complexity to be addressed through comparisons of model fit based on modern information criteria, such as Pareto smoothed importance sampling approximate leave-one-out cross validation [$\text{PSIS-LOO}$; @gelman.etal2014; @vehtari.etal2017].[^2] Second, the Bayesian framework supports straightforward implementation of hierarchical versions of these models allowing for information about model parameters to be shared across participants, resulting in potential improvements to out-of-sample prediction, reductions in model complexity, and a more realistic test of the models. Finally, a Bayesian framework also supports new extensions of these models to directly model participants trial-level responses while accounting for rounding and out-of-distribution response errors. These extensions allow for principled probabilistic tests of the distributional predictions of the models.

@@

@@

This first set of analyses compares the models' ability to capture participant's probability judgments in expectation, averaged over the three blocks on which they made judgments about each query. These analyses will test the first points of differentiation between the models: their different predictions with respect to conditional probability judgments and their specific parameterizations.

@@

Both of these models can also be implemented as hierarchical models with partial pooling for the $d$ and $d'$ parameters (implicitly, for $N$ and $N'$ in the case of the Bayesian Sampler). This partial pooling can help to regularize parameter estimates and improve out-of-sample predictive performance. In addition, partial pooling effectively reduces model complexity, and could support more realistic comparison between the "simple" and "complex" variants of the models. Figure 2 displays the translation of a hierarchical implementation of the Bayesian Sampler model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters. For ease of interpretation, the centered parameterization is shown below, although the actual models used a non-centered parameterization to improve sampling efficiency [@papaspiliopoulos.etal2007].

@@

Finally, I also explored fitting a hierarchical version of the Bayesian Sampler model that allowed values of $\beta > 1$. Restricting $\beta$ to [0,1] restricts the prior distribution of the Bayesian sampler to the class of "ignorance priors" [@zhu.etal2020]. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted [see e.g., @fennell.baddeley2012]. If $\beta$ is unrestricted, allowed to fall in the domain $[0, \infty]$ then the Bayesian Sampler model becomes more flexible, allowing for equivalent "noise" levels in the same $[0, 1/2]$ range as the PT+N model. That is, through the bridging condition, the implied $d$ approaches $1/2$ in the limit as $N \to 1$ and $\beta \to \infty$. Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its two-stage process of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

Data from Experiment 1 favor "complex" variants of the Bayesian Sampler model compared with the "simple" variants and all versions of the PT+N model (greater values of $\widehat{\text{elpd}}_{\text{LOO}}$ are better). As shown in Figure 3, the best-scoring model is an unrestricted variant of the Bayesian Sampler that allows for people to bring informative priors to the probability judgment task (i.e. allowing $\beta \in [0, \infty]$. Data from Experiment 2 more decisively reveal a single winning model: the hierarchical "unrestricted" implementation of the Bayesian Sampler model allowing for informative priors.

This unrestricted BS model differs from the PT+N model only in its treatment of conditional probability judgments and so, from its superior fit, we can infer that the Bayesian Sample theory provides a better account of human conditional probability judgments.

@@

```{r, fig.cap="Posterior density of population-level $d$  and $d'$ parameters estimated from the unrestricted hierarchical Bayesian Sampler model for data from Experiments 1 and 2. Dashed line indicates theoretical maximum values for Bayesian Sampler model with uninformative priors."}
plt_d_avgmodel + plt_d_trialmodel + plot_layout(ncol=1, guides="collect") & labs(y = "Density", x="Estimate") & theme_bw() & theme(panel.grid=element_blank())

@@

t2_df %>% 
  select(-`...1`) %>% 
  gather(measure, value, -Experiment, -model) %>% 
  pivot_wider(names_from = c(Experiment, measure), values_from=value,names_sort = T) %>% 
  rename(Model = model) %>%
  kbl(
    caption="Bayesian model comparison results with best scoring model in bold face. Models are compared based on $\\widehat{\text{elpd}}_{\text{LOO}}$ and the correlation between their predictions for each participants' query-level average ($r_{\\text{query}}$) as well as overall query-level averages $r_{\\text{query-avg}}$.", 
    digits=3, 
    booktabs=T,
    col.names = c("Model", "$\\widehat{\\text{elpd}}_{\\text{LOO}}$", "$\\hat{p}_{\\text{LOO}}$","$r_{\\text{query}}$", "$r_{\\text{query-avg}}$", "$\\widehat{\\text{elpd}}_{\\text{LOO}}$", "$\\hat{p}_{\\text{LOO}}$","$r_{\\text{query}}$", "$r_{\\text{query-avg}}$"),
    escape = F
    ) %>%
  add_header_above(c(" ", "Experiment 1"=4, "Experiment 2"=4)) %>% 
  kable_classic(full_width=F) %>% 
  kable_styling(latex_options="scale_down") %>% 
  row_spec(1, bold=TRUE)
```

@@

```{r, fig.cap="Average model predicted and observed values for the 18 identities. Note that the Bayesian Sampler but not the PT+N model is capable of predicting non-zero values for identities Z10 through Z13. Error bars represent 95% CI. In Experiment 1, like participants' responses, the model's estimates are very slightly positive for \\{icy, frosty\\} and very slightly negative for \\{normal, typical\\}. This pattern replicates the qualitative pattern reported by Zhu and colleagues."}
plt_ineq_exp1 + labs(title="Exp. 1") + 
  plt_ineq_exp2 + labs(title="Exp. 2") + 
  plot_layout(guides="collect") & 
  theme(legend.position="bottom", axis.text.x = element_text(angle=90, hjust=1, vjust=.5)) & 
  labs(color="", y = "Estimate", x = "Identity")
```

@@

## Results: Raw trial-level response distributions

 The best-fitting model capturing participant's predictions in expectation can be seen either as a variant of the Bayesian Sampler theory allowing for informative priors or as a variant of the PT+N theory without any special treatment of conditional versus unconditional probability judgments. Thus, these prior analyses have not decisively ruled between the different process-level psychological theories behind the models. To evaluate the Bayesian Sampler and PT+N theories' competing accounts of the psychological processes behind probability judgments, a second set of analyses focused on the distribution of participant's trial-by-trial responses was conducted.

Recall, in contrast to a noise-based account, the Bayesian Sampler theory predicts both a truncation of the response distribution as well as a correlation between the degree of shrinkage in probability estimates and the variability of those estimates trial-to-trial (assuming $\beta$ is relatively small). Because the Bayesian Sampler applies a Bayesian adjustment after sampling, it predicts probability judgments will always lie between $d$ and $1 - d$, even when zero positive or negative mental samples are drawn (see Equation 12 and the bridging condition, Equation 5). First, it bears noting that the truncated response distribution implied by the Bayesian Sampler model appears at odds with the raw response data: participants' responses frequently lie outside the range implied by the best estimates of their $d$ parameters (41% in Experiment 1 and 60% in Experiment 2).

@@

Under both models, the variability of people's responses trial-to-trial is driven by the number of mental samples drawn: more mental samples produce less-variable responses. However, if the number of samples is considered to be a truly discrete quantity, then only a limited number of discrete responses are possible. As Zhu and colleauges [-@zhu.etal2020] note, this is somewhat implausible on its face and their later work has abandoned this assumption [@zhu.etal2021]. At the same time, from a pragmatic perspective it is highly desirable that all latent parameters within the models be continuous rather than discrete. The models would be far more tractable to fit if, rather than including a latent Binomial variable representing the discrete number of samples drawn, we could instead model a continuous proportion of samples using, for instance, a Beta distribution.

Zhu and collauges [-@zhu.etal2020] introduce the possibility of an "autocorrelated" Bayesian Sampler model under which samples are assumed to be autocorrelated [ideas which were advanced further in @zhu.etal2021]. As autocorrelated samples provide less information than i.i.d samples, they should be weighted when computing probability estimates. The idea is that people actually draw N autocorrelated samples that approximate some smaller effective number of i.i.d samples ($N_{\text{eff}}$). Assuming the actual number of autocorrelated samples drawn, $N$, is allowed to vary somewhat noisily, then a model based on autocorrelated samples would no longer be limited to predicting a discrete set of possible responses. To approximate this as part of a wholly continuous model, I model the proportions calculated from such a hypothetical autocorrelated sampling process using a Beta distribution.

@@

First, it is clear that participants have rounded a majority of their responses. This sort of rounding can be modeled by a categorical distribution across the discrete possible rounded responses. Each rounded response category corresponds to a set of cut points, $a$ and $b$, with the probability of the categorical response defined by the cumulative distribution function of the underlying latent distribution ($B$).

@@

As participants were allowed to respond freely with whole numbers from 0 to 100, the exact rounding policy for each response is unknown. Nevertheless these rounding policies can be estimated via mixture modeling. For simplicity, rounding to the nearest 5% was enforced for all responses. Then, the probability of these categorical responses are computed for 21 and 11 categories (corresponding to rounding to 5% and 10%). These probabilities were combined along with a uniform probability representing "contaminants" according to mixing probabilities $\phi$, distributed with a Dirichlet prior (see Appendix for further implementation details).

@@

Approximating an autocorrelated sampling process using a Beta distribution and starting from the original Bayesian Sampler model,

@@

This model assumes that the number of samples drawn on each trial is fixed as $N$ or $N'$ accordingly (modulo the uncertainty about these parameters). However, it also seems reasonable to imagine that the number of samples drawn in fact varies from trial-to-trial. For the Bayesian Sampler model, this could substantially impact the model's fit. As the number of samples drawn affects the truncation of the response distribution, this may allow the model to capture some responses that would otherwise be treated as contaminants. To capture this, the model can be given one additional extension to allow the number of samples to vary, by adding a new parameter $N_{trial}$. This parameter multiplies the number of effective samples as a fraction of each individual participants' average number of samples drawn, e.g. so that a participant might sometimes draw 1.5x or 2x the number of effective samples they typically draw. The appropriate amount of variation in $N_{trial}$ is constrained to be fairly small, but is estimated hierarchically: I assume $log(N_{trial}) \sim N(0,\sigma_{N_{trial}})$ and $log(\sigma_{N_{trial}}) \sim (-1, .3)$.

@@

Prior to fitting the models, response data were rounded to the nearest 5%. Nearly all responses (Exp. 1: 93% and Exp. 2: 89%) were already divisible by five, and this was necessary to speed model fitting. Even still, estimating model posteriors for the trial-level mixture models using MCMC proved intractable. Instead, model posteriors were estimated using Stochastic Variational Inference (SVI) with Numpyro [@phan.etal2019] using a multivariate Normal guide [e.g. @kucukelbir.etal2015]. Estimating each model posterior took between approximately 20 to 90 minutes on an Nvidia V100 GPU. Simulation studies verified that this estimation approach could reliably recover parameters from simulated data (see Supplemental Materials).

@@

The non-varying Bayesian Sampler model fits quite poorly and is estimated to have a large proportion of "contaminant" responses (43% in Experiment 1, 26% in Experiment 2). This is likely due to a lack of the predicted truncation of the response distribution. As with the trial-average models, the estimates of implied d are quite high, which would predict substantial truncation. As mentioned earlier, many of participants' responses fall outside the range implied by their most likely implied d values.

Allowing the effective number of samples drawn to vary trial-by-trial improves the fit of the Bayesian Sampler model substantially and somewhat decreases the estimated proportion of contaminant responses (30% in Experiment 1, 28% in Experiment 2). Nevertheless, these results still indicate inferior fit compared with the noise-based model, which has substantially lower $\widehat{\text{elpd}}_{\text{LOO}}$ and attributes far fewer responses to the contaminant process (11% in Experiment 1, 4% in Experiment 2).

Second, we can examine the shrinkage-variance relationship. In the noise-based model, $N$ and $d$ were allowed to vary freely. But if these quantities are actually correlated as predicted by a Bayesian adjustment model, then we should expect to nevertheless see a correlation between the subject-level estimates in these parameters. Figure 11 shows scatterplots relating these estimates. Although there is a slight negative correlation in Experiment 1, this is driven by only a handful of participants with extreme values. In Experiment 2 the correlation appears to if anything be positive rather than negative. These findings again run counter to the predictions of the Bayesian Sampler theory.

```{r, fig.cap = "Scatterplots showing the relationships between subject-level d and N estimates from the trial-level noise-based model for Experiments 1 and 2. Figure for Experiment 1 excludes some outlier paricipants who gave repetitive responses that resulted in abnormally high N-values. Error bars indicate 95% CI. "}
